### 2024-10-10 manuscript.tex
add draft for abstract [ci skip]
This build is based on
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/876267bf5f5cb653b6898a4b032be3be7ca3b9f8.

This commit was created by the following CI build and job:
https://github.com/pivlab/annual_review_of_biomedical_data_science/commit/876267bf5f5cb653b6898a4b032be3be7ca3b9f8/checks
https://github.com/pivlab/annual_review_of_biomedical_data_science/actions/runs/11282708179

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/517d43c379923fb79c0ec18f4a2aa10ed2beb727/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/517d43c379923fb79c0ec18f4a2aa10ed2beb727}{pivlab/annual\_review\_of\_biomedical\_data\_science@517d43c}
on October 10, 2024.

@@

This is a very cool paper about something.
We are doing this and that.
And this are the conclusions.
And this is in \textbf{boldface}.

@@

Approaches based on gene modules have the potential to go beyond prior knowledge and capture patterns that might be unique to different disease datasets.
First, although a gene module can be aligned to a pathway, this does not mean that it is restricted to prior knowledge as in standard pathway analyses.
Being aligned with pathways means that a module resembles a known mechanism, and this helps in interpretability and also in separating from other patterns that might be related to technical noise but the module could capture other genes as well that are not part of the pathway but potentially also involved the function.
Second, gene modules can also be extracted from very large, heterogeneous datasets such as recount2, or more specific ones such as the Human Trisome Project, which includes gene expression from people with Down Syndrome (DS).
Approaches such as PLIER and PhenoPLIER have been fundamental to extract DS-specific gene modules related to obesity, a common co-occurring condition The omni-PLIER framework, combining the PLIER model with the omnigenic approach, uncovers molecular mechanisms linking Down syndrome (DS) and obesity, identifying key latent variables with top genes such as GPX1 and MCL1, which mediate the relationship between trisomy 21 and body mass index (BMI), offering insights into DS-specific and general obesity-related processes. \texttt{{[}CITE\ Sutanu’s\ paper{]}}.

A disadvantage of the PhenoPLIER approach is that gene modules are generated by an algorithm, and as such, they could represent artifacts or be aligned with technical noise.
Interpretable methods such as PLIER and some VAE models that help segregate technical noise from relevant biology are key in solving this problem.
Another limitation is the potential difficulty in determining the contexts in which a gene module is expressed.
In the example of LV256 and lipid-related traits, the module was also found to be expressed, with different degrees, in tumor samples, skin, skeletal muscle and other contexts that might be hard to interpret.
The interpretability advantage is also limited by the quality of the RNA-seq metadata.
Additionally, compared with GWAS and TWAS, where the unit is an objective molecular entity (a SNP or gene in a particular chromosome and position), a gene module not only might or might not exist, but the algorithm could fail to capture all gene modules relevant for a particular study, which could bias any downstream analysis.
Finally, a gene module, equivalent to a gene co-expression network, represents a set of correlated genes' expression in some conditions, but it does not provide information about gene-gene links (it's not directed) and could include many false positives {[}\protect\hyperlink{ref-jVQ2rMqc}{94}{]}.
Future approaches should also incorporate other data modalities to refine a gene module such as transcription factor binding data or chromatin accessibility.

Although there are several limitations when we only correlate gene pairs across RNA-seq samples {[}\protect\hyperlink{ref-jVQ2rMqc}{94}{]}, integrating gene networks with other data modalities have the potential to identify disease-relevant molecular mechanisms {[}\protect\hyperlink{ref-dg9nKuy0}{54}{]}.
Here we described an approach that integrates gene modules with other data sources such as gene-trait and gene-drug information.
The underlying hypothesis of this approach is that biologically meaningful gene-gene links will be captured when different data modalities that at least partially capture these links are integrated.
In our examples, these data modalities are gene modules from transcriptome data, single gene-trait associations and single gene-drug links via drug-induced transcriptional profiles.
We also showed that if the unsupervised approach used to extract a gene module is also interpretable (i.e., provides information about the specific contexts such as cell types or tissues that explain why those genes were grouped together), the data integration approach also sheds light on potential context-specific transcriptional mechanisms.

@@

Genomics plays an important role in realizing the promise of personalized medicine, although not without significant challenges.
In this review, we discuss different molecular phenotypes, theoretical models and computational approaches that offer different mechanistic views on how genetic variation leads to phenotypic changes.
To fully understand the molecular basis of human complex traits, it is essential to integrate diverse datasets and unravel the complex structures of biological networks.
Gene co-expression networks, represented as gene modules, have been successfully used across a variety of fields {[}\protect\hyperlink{ref-CqvNaEqc}{95},\protect\hyperlink{ref-r8alvfcC}{96},\protect\hyperlink{ref-E1tpJnqG}{97}{]}.
Here, we reviewed the main components of a gene module-based approach that leverages gene co-expression patterns learned from large transcriptomic datasets to integrate GWAS, TWAS, and drug-induced transcriptional profiles.
Drawing on modern theories of the genetic architecture of complex traits, we show that these approaches can prioritize core genes and create interpretable frameworks for drug repurposing.

Gene-gene interactions are a key component in the omnigenic model.
While the integration of gene modules with gene-trait associations and gene-drug data has been a valuable first step, it represents only a fraction of the complexity inherent in biological systems.
Biological networks extend far beyond these initial layers, involving intricate, multilayered systems such as histone modifications, transcription factor binding sites, \emph{cis}-regulatory networks, and chromatin conformation---all of which contribute to the co-regulation of gene transcription.
Moreover, this complexity includes protein-protein interaction networks, which form the backbone of signaling pathways and facilitate the transmission of signals both within and between cells.
Additionally, metabolic networks influence enzyme activities and metabolite levels, propagating changes that affect numerous other molecules {[}\protect\hyperlink{ref-dg9nKuy0}{54},\protect\hyperlink{ref-KNAOTAKP}{98}{]}.
Incorporating and quantifying these additional biological layers will enhance gene module construction, offering a more comprehensive and accurate representation of the intricate nature of real biological networks.

A gene module approach that integrates genetic studies has several important inputs: 1) gene expression data relevant to the research problem, such as large heterogeneous datasets that span over different conditions, or otherwise focused, smaller dataset that capture more subtle, unique, and disease-specific gene modules; 2) unbiased and relevant prior information that can be used during or after module extraction to distinguish relevant transcriptomic signatures from technical noise; and 3) reliable single variant/gene associations with the complex traits of interest.
Therefore, it is important to emphasize that a gene module approach is by no means a replacement of current single variant/gene approaches.
The development of new and improved statistical approaches that detect more causal genes, regardless of whether they have a direct or indirect effect on the trait, while reducing false positives {[}\protect\hyperlink{ref-i8Qjy4pw}{99}{]} will only improve other approaches that rely on them, such as gene module-based methods described here.
In this review, we show that the identification of peripheral genes is key to prioritize core genes that might represent more attractive drug targets, strongly suggesting that the mechanism proposed by the omnigenic model, i.e., that gene regulatory networks are highly interconnected and genes have specific roles with more direct or indirect effects on traits, is useful.

@@

Leveraging GPU acceleration holds great promise for significantly speeding up bioinformatics workflows.
Research has shown that GPU technology can provide remarkable performance boosts, with speed improvements reaching up to 1000x for specific tasks {[}\protect\hyperlink{ref-1EIbaYHGq}{100}{]}.
Integrating GPU-aware libraries like CuPy {[}\protect\hyperlink{ref-JMQXR7X9}{101}{]} and RAPIDS {[}\protect\hyperlink{ref-wPsNXL1Q}{102}/{]} into our tools allows for seamless performance upgrades in established frameworks such as NumPy {[}\protect\hyperlink{ref-1LexfAxj}{103}{]} and Scikit-learn {[}\protect\hyperlink{ref-AujvwLp6}{104}{]}.
As our tools evolve, we anticipate developing custom GPU kernels, further enhancing speed and optimizing them for specific computational needs.
Advanced GPU programming with PyCUDA {[}\protect\hyperlink{ref-b5xmRoLZ}{105}{]} could also be key to future optimizations.

Another promising direction is utilizing WebAssembly {[}\protect\hyperlink{ref-ITQaZvrQ}{106}{]} to make bioinformatics tools more portable and accessible.
We aim to adapt existing bioinformatics applications into web-based tools that can be run directly in the browser, providing users with full control over their data and eliminating the need for data transfer over networks.
Projects like Biowasm {[}\protect\hyperlink{ref-Z0f406Zs}{107}/{]}, Kana {[}\protect\hyperlink{ref-1DswCJYT0}{108}{]}, and VirtualWasm {[}\protect\hyperlink{ref-WhJVE7a4}{109}{]} have already demonstrated the potential of WebAssembly for genomics and single-cell analysis, and we plan to build on this foundation.
By executing statistical analyses, computations, and visualizations directly on the user's device, WebAssembly will not only address privacy and compliance concerns but also reduce server costs and computational load by shifting the workload to the client side.

@@

\leavevmode\vadjust pre{\hypertarget{ref-jVQ2rMqc}{}}%
\CSLLeftMargin{94. }%
\CSLRightInline{\textbf{Gene regulatory network inference in the era of single-cell multi-omics}
\CSLBlock{Pau Badia-i-Mompel, Lorna Wessels, Sophia Müller-Dott, Rémi Trimbour, Ricardo O Ramirez Flores, Ricard Argelaguet, Julio Saez-Rodriguez} \emph{Nature Reviews Genetics} (2023-06-26) \url{https://doi.org/gsdq3p}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41576-023-00618-5}{10.1038/s41576-023-00618-5} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/37365273}{37365273}}}

\leavevmode\vadjust pre{\hypertarget{ref-CqvNaEqc}{}}%
\CSLLeftMargin{95. }%
\CSLRightInline{\textbf{WGCNA: an R package for weighted correlation network analysis}
\CSLBlock{Peter Langfelder, Steve Horvath} \emph{BMC Bioinformatics} (2008-12) \url{https://doi.org/bwdbzc}
\CSLBlock{DOI: \href{https://doi.org/10.1186/1471-2105-9-559}{10.1186/1471-2105-9-559} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/19114008}{19114008} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2631488}{PMC2631488}}}

\leavevmode\vadjust pre{\hypertarget{ref-r8alvfcC}{}}%
\CSLLeftMargin{96. }%
\CSLRightInline{\textbf{WGCNA Application to Proteomic and Metabolomic Data Analysis}
\CSLBlock{G Pei, L Chen, W Zhang} \emph{Methods in Enzymology} (2017) \url{https://doi.org/grksc8}
\CSLBlock{DOI: \href{https://doi.org/10.1016/bs.mie.2016.09.016}{10.1016/bs.mie.2016.09.016} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/28109426}{28109426}}}

\leavevmode\vadjust pre{\hypertarget{ref-E1tpJnqG}{}}%
\CSLLeftMargin{97. }%
\CSLRightInline{\textbf{Identification of glioblastoma~gene prognosis modules based on weighted gene co-expression network analysis}
\CSLBlock{Pengfei Xu, Jian Yang, Junhui Liu, Xue Yang, Jianming Liao, Fanen Yuan, Yang Xu, Baohui Liu, Qianxue Chen} \emph{BMC Medical Genomics} (2018-11-01) \url{https://doi.org/gfzt4j}
\CSLBlock{DOI: \href{https://doi.org/10.1186/s12920-018-0407-1}{10.1186/s12920-018-0407-1} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/30382873}{30382873} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6211550}{PMC6211550}}}

\leavevmode\vadjust pre{\hypertarget{ref-KNAOTAKP}{}}%
\CSLLeftMargin{98. }%
\CSLRightInline{\textbf{Boosting GWAS using biological networks: A study on susceptibility to familial breast cancer}
\CSLBlock{Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott} \emph{PLOS Computational Biology} (2021-03-18) \url{https://doi.org/g7zdns}
\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pcbi.1008819}{10.1371/journal.pcbi.1008819} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/33735170}{33735170} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8009366}{PMC8009366}}}

\leavevmode\vadjust pre{\hypertarget{ref-i8Qjy4pw}{}}%
\CSLLeftMargin{99. }%
\CSLRightInline{\textbf{Adjusting for genetic confounders in transcriptome-wide association studies improves discovery of risk genes of complex traits}
\CSLBlock{Siming Zhao, Wesley Crouse, Sheng Qian, Kaixuan Luo, Matthew Stephens, Xin He} \emph{Nature Genetics} (2024-01-26) \url{https://doi.org/gtf7kb}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41588-023-01648-9}{10.1038/s41588-023-01648-9} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/38279041}{38279041} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10864181}{PMC10864181}}}

\leavevmode\vadjust pre{\hypertarget{ref-1EIbaYHGq}{}}%
\CSLLeftMargin{100. }%
\CSLRightInline{\textbf{Graphics processing units in bioinformatics, computational biology and systems biology}
\CSLBlock{Marco S Nobile, Paolo Cazzaniga, Andrea Tangherloni, Daniela Besozzi} \emph{Briefings in Bioinformatics} (2016-07-08) \url{https://doi.org/g7zg6w}
\CSLBlock{DOI: \href{https://doi.org/10.1093/bib/bbw058}{10.1093/bib/bbw058} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/27402792}{27402792} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5862309}{PMC5862309}}}

\leavevmode\vadjust pre{\hypertarget{ref-JMQXR7X9}{}}%
\CSLLeftMargin{101. }%
\CSLRightInline{\textbf{cupy/cupy}
\CSLBlock{CuPy} (2024-10-10) \url{https://github.com/cupy/cupy}}

\leavevmode\vadjust pre{\hypertarget{ref-wPsNXL1Q}{}}%
\CSLLeftMargin{102. }%
\CSLRightInline{\textbf{RAPIDS \textbar{} GPU Accelerated Data Science}
\CSLBlock{RAPIDS \textbar{} GPU Accelerated Data Science} \url{https://rapids.ai/}}

\leavevmode\vadjust pre{\hypertarget{ref-1LexfAxj}{}}%
\CSLLeftMargin{103. }%
\CSLRightInline{\textbf{Array programming with NumPy}
\CSLBlock{Charles R Harris, KJarrod Millman, Stéfan J van der Walt, Ralf Gommers, Pauli Virtanen, David Cournapeau, Eric Wieser, Julian Taylor, Sebastian Berg, Nathaniel J Smith, \ldots{} Travis E Oliphant} \emph{Nature} (2020-09-16) \url{https://doi.org/ghbzf2}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41586-020-2649-2}{10.1038/s41586-020-2649-2} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/32939066}{32939066} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7759461}{PMC7759461}}}

\leavevmode\vadjust pre{\hypertarget{ref-AujvwLp6}{}}%
\CSLLeftMargin{104. }%
\CSLRightInline{\textbf{Scikit-learn: Machine Learning in Python}
\CSLBlock{Fabian Pedregosa, Gaël Varoquaux, Alexandre Gramfort, Vincent Michel, Bertrand Thirion, Olivier Grisel, Mathieu Blondel, Andreas Müller, Joel Nothman, Gilles Louppe, \ldots{} Édouard Duchesnay} \emph{arXiv} (2012) \url{https://doi.org/gpptwn}
\CSLBlock{DOI: \href{https://doi.org/10.48550/arxiv.1201.0490}{10.48550/arxiv.1201.0490}}}

\leavevmode\vadjust pre{\hypertarget{ref-b5xmRoLZ}{}}%
\CSLLeftMargin{105. }%
\CSLRightInline{\textbf{PyCUDA and PyOpenCL: A scripting-based approach to GPU run-time code generation}
\CSLBlock{Andreas Klöckner, Nicolas Pinto, Yunsup Lee, Bryan Catanzaro, Paul Ivanov, Ahmed Fasih} \emph{Parallel Computing} (2012-03) \url{https://doi.org/c8hxzm}
\CSLBlock{DOI: \href{https://doi.org/10.1016/j.parco.2011.09.001}{10.1016/j.parco.2011.09.001}}}

>>>>

This manuscript
(\href{https://pivlab.github.io/annual_review_of_biomedical_data_science/v/876267bf5f5cb653b6898a4b032be3be7ca3b9f8/}{permalink})
was automatically generated
from \href{https://github.com/pivlab/annual_review_of_biomedical_data_science/tree/876267bf5f5cb653b6898a4b032be3be7ca3b9f8}{pivlab/annual\_review\_of\_biomedical\_data\_science@876267b}
on October 10, 2024.

@@

Understanding the genetic basis of complex traits is a longstanding challenge in the field of genomics.
Genome-wide association studies (GWAS) have identified thousands of variant-trait associations, but most of these variants are located in non-coding regions, making the link to biological function elusive.
While traditional approaches, such as Transcriptome-wide association studies (TWAS), have advanced our understanding by linking genetic variants to gene expression, they often overlook gene-gene interactions.
Here, we review current approaches based on gene modules to integrate different molecular data, leveraging machine learning methods to identify gene modules based on co-expression and functional relationships.
These integrative approaches, like PhenoPLIER, combine TWAS and drug-induced transcriptional profiles to effectively capture biologically meaningful gene networks.
This integration provides a context-specific understanding of disease proceses while highlighting both core and peripheral genes.
These insights pave the way for novel therapeutic targets and enhance the interpretability of genetic studies in personalized medicine.

@@

Approaches based on gene modules have the potential to go beyond prior knowledge and capture patterns that might be unique to different disease datasets.
First, although a gene module can be aligned to a pathway, this does not mean that it is restricted to prior knowledge as in standard pathway analyses.
Being aligned with pathways means that a module resembles a known mechanism, and this helps in interpretability and also in separating from other patterns that might be related to technical noise but the module could capture other genes as well that are not part of the pathway but potentially also involved the function.
Second, gene modules can also be extracted from very large, heterogeneous datasets such as recount2, or more specific ones such as the Human Trisome Project, which includes gene expression from people with Down Syndrome (DS).
Approaches such as PLIER and PhenoPLIER have been fundamental to extract DS-specific gene modules related to obesity, a common co-occurring condition The omni-PLIER framework, combining the PLIER model with the omnigenic approach, uncovers molecular mechanisms linking Down syndrome (DS) and obesity, identifying key latent variables with top genes such as GPX1 and MCL1, which mediate the relationship between trisomy 21 and body mass index (BMI), offering insights into DS-specific and general obesity-related processes {[}\protect\hyperlink{ref-B7FF0NeZ}{94}{]}.

A disadvantage of the PhenoPLIER approach is that gene modules are generated by an algorithm, and as such, they could represent artifacts or be aligned with technical noise.
Interpretable methods such as PLIER and some VAE models that help segregate technical noise from relevant biology are key in solving this problem.
Another limitation is the potential difficulty in determining the contexts in which a gene module is expressed.
In the example of LV256 and lipid-related traits, the module was also found to be expressed, with different degrees, in tumor samples, skin, skeletal muscle and other contexts that might be hard to interpret.
The interpretability advantage is also limited by the quality of the RNA-seq metadata.
Additionally, compared with GWAS and TWAS, where the unit is an objective molecular entity (a SNP or gene in a particular chromosome and position), a gene module not only might or might not exist, but the algorithm could fail to capture all gene modules relevant for a particular study, which could bias any downstream analysis.
Finally, a gene module, equivalent to a gene co-expression network, represents a set of correlated genes' expression in some conditions, but it does not provide information about gene-gene links (it's not directed) and could include many false positives {[}\protect\hyperlink{ref-jVQ2rMqc}{95}{]}.
Future approaches should also incorporate other data modalities to refine a gene module such as transcription factor binding data or chromatin accessibility.

Although there are several limitations when we only correlate gene pairs across RNA-seq samples {[}\protect\hyperlink{ref-jVQ2rMqc}{95}{]}, integrating gene networks with other data modalities have the potential to identify disease-relevant molecular mechanisms {[}\protect\hyperlink{ref-dg9nKuy0}{54}{]}.
Here we described an approach that integrates gene modules with other data sources such as gene-trait and gene-drug information.
The underlying hypothesis of this approach is that biologically meaningful gene-gene links will be captured when different data modalities that at least partially capture these links are integrated.
In our examples, these data modalities are gene modules from transcriptome data, single gene-trait associations and single gene-drug links via drug-induced transcriptional profiles.
We also showed that if the unsupervised approach used to extract a gene module is also interpretable (i.e., provides information about the specific contexts such as cell types or tissues that explain why those genes were grouped together), the data integration approach also sheds light on potential context-specific transcriptional mechanisms.

@@

Genomics plays an important role in realizing the promise of personalized medicine, although not without significant challenges.
In this review, we discuss different molecular phenotypes, theoretical models and computational approaches that offer different mechanistic views on how genetic variation leads to phenotypic changes.
To fully understand the molecular basis of human complex traits, it is essential to integrate diverse datasets and unravel the complex structures of biological networks.
Gene co-expression networks, represented as gene modules, have been successfully used across a variety of fields {[}\protect\hyperlink{ref-CqvNaEqc}{96},\protect\hyperlink{ref-r8alvfcC}{97},\protect\hyperlink{ref-E1tpJnqG}{98}{]}.
Here, we reviewed the main components of a gene module-based approach that leverages gene co-expression patterns learned from large transcriptomic datasets to integrate GWAS, TWAS, and drug-induced transcriptional profiles.
Drawing on modern theories of the genetic architecture of complex traits, we show that these approaches can prioritize core genes and create interpretable frameworks for drug repurposing.

Gene-gene interactions are a key component in the omnigenic model.
While the integration of gene modules with gene-trait associations and gene-drug data has been a valuable first step, it represents only a fraction of the complexity inherent in biological systems.
Biological networks extend far beyond these initial layers, involving intricate, multilayered systems such as histone modifications, transcription factor binding sites, \emph{cis}-regulatory networks, and chromatin conformation---all of which contribute to the co-regulation of gene transcription.
Moreover, this complexity includes protein-protein interaction networks, which form the backbone of signaling pathways and facilitate the transmission of signals both within and between cells.
Additionally, metabolic networks influence enzyme activities and metabolite levels, propagating changes that affect numerous other molecules {[}\protect\hyperlink{ref-dg9nKuy0}{54},\protect\hyperlink{ref-KNAOTAKP}{99}{]}.
Incorporating and quantifying these additional biological layers will enhance gene module construction, offering a more comprehensive and accurate representation of the intricate nature of real biological networks.

A gene module approach that integrates genetic studies has several important inputs: 1) gene expression data relevant to the research problem, such as large heterogeneous datasets that span over different conditions, or otherwise focused, smaller dataset that capture more subtle, unique, and disease-specific gene modules; 2) unbiased and relevant prior information that can be used during or after module extraction to distinguish relevant transcriptomic signatures from technical noise; and 3) reliable single variant/gene associations with the complex traits of interest.
Therefore, it is important to emphasize that a gene module approach is by no means a replacement of current single variant/gene approaches.
The development of new and improved statistical approaches that detect more causal genes, regardless of whether they have a direct or indirect effect on the trait, while reducing false positives {[}\protect\hyperlink{ref-i8Qjy4pw}{100}{]} will only improve other approaches that rely on them, such as gene module-based methods described here.
In this review, we show that the identification of peripheral genes is key to prioritize core genes that might represent more attractive drug targets, strongly suggesting that the mechanism proposed by the omnigenic model, i.e., that gene regulatory networks are highly interconnected and genes have specific roles with more direct or indirect effects on traits, is useful.

@@

Leveraging GPU acceleration holds great promise for significantly speeding up bioinformatics workflows.
Research has shown that GPU technology can provide remarkable performance boosts, with speed improvements reaching up to 1000x for specific tasks {[}\protect\hyperlink{ref-1EIbaYHGq}{101}{]}.
Integrating GPU-aware libraries like CuPy {[}\protect\hyperlink{ref-4iZVPVHz}{102}{]} and RAPIDS {[}\protect\hyperlink{ref-wPsNXL1Q}{103}/{]} into our tools allows for seamless performance upgrades in established frameworks such as NumPy {[}\protect\hyperlink{ref-1LexfAxj}{104}{]} and Scikit-learn {[}\protect\hyperlink{ref-AujvwLp6}{105}{]}.
As our tools evolve, we anticipate developing custom GPU kernels, further enhancing speed and optimizing them for specific computational needs.
Advanced GPU programming with PyCUDA {[}\protect\hyperlink{ref-b5xmRoLZ}{106}{]} could also be key to future optimizations.

Another promising direction is utilizing WebAssembly {[}\protect\hyperlink{ref-ITQaZvrQ}{107}{]} to make bioinformatics tools more portable and accessible.
We aim to adapt existing bioinformatics applications into web-based tools that can be run directly in the browser, providing users with full control over their data and eliminating the need for data transfer over networks.
Projects like Biowasm {[}\protect\hyperlink{ref-Z0f406Zs}{108}/{]}, Kana {[}\protect\hyperlink{ref-1DswCJYT0}{109}{]}, and VirtualWasm {[}\protect\hyperlink{ref-WhJVE7a4}{110}{]} have already demonstrated the potential of WebAssembly for genomics and single-cell analysis, and we plan to build on this foundation.
By executing statistical analyses, computations, and visualizations directly on the user's device, WebAssembly will not only address privacy and compliance concerns but also reduce server costs and computational load by shifting the workload to the client side.

@@

\leavevmode\vadjust pre{\hypertarget{ref-B7FF0NeZ}{}}%
\CSLLeftMargin{94. }%
\CSLRightInline{\textbf{A Pathway-Level Information Extractor (PLIER) Framework to Gain Mechanistic Insights into Obesity in Down Syndrome}
\CSLBlock{Sutanu Nandi, Yuehua Zhu, Lucas A Gillenwater, Marc Subirana Ganes, Haoyu Zhang, Negar Janani, Casey Greene, Milton Pividori, Maria Chikina, James C Costello} \emph{Proceedings of the Pacific Symposium on Biocomputing 2025 (PSB2025)} (2024)}

\leavevmode\vadjust pre{\hypertarget{ref-jVQ2rMqc}{}}%
\CSLLeftMargin{95. }%
\CSLRightInline{\textbf{Gene regulatory network inference in the era of single-cell multi-omics}
\CSLBlock{Pau Badia-i-Mompel, Lorna Wessels, Sophia Müller-Dott, Rémi Trimbour, Ricardo O Ramirez Flores, Ricard Argelaguet, Julio Saez-Rodriguez} \emph{Nature Reviews Genetics} (2023-06-26) \url{https://doi.org/gsdq3p}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41576-023-00618-5}{10.1038/s41576-023-00618-5} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/37365273}{37365273}}}

\leavevmode\vadjust pre{\hypertarget{ref-CqvNaEqc}{}}%
\CSLLeftMargin{96. }%
\CSLRightInline{\textbf{WGCNA: an R package for weighted correlation network analysis}
\CSLBlock{Peter Langfelder, Steve Horvath} \emph{BMC Bioinformatics} (2008-12) \url{https://doi.org/bwdbzc}
\CSLBlock{DOI: \href{https://doi.org/10.1186/1471-2105-9-559}{10.1186/1471-2105-9-559} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/19114008}{19114008} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2631488}{PMC2631488}}}

\leavevmode\vadjust pre{\hypertarget{ref-r8alvfcC}{}}%
\CSLLeftMargin{97. }%
\CSLRightInline{\textbf{WGCNA Application to Proteomic and Metabolomic Data Analysis}
\CSLBlock{G Pei, L Chen, W Zhang} \emph{Methods in Enzymology} (2017) \url{https://doi.org/grksc8}
\CSLBlock{DOI: \href{https://doi.org/10.1016/bs.mie.2016.09.016}{10.1016/bs.mie.2016.09.016} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/28109426}{28109426}}}

\leavevmode\vadjust pre{\hypertarget{ref-E1tpJnqG}{}}%
\CSLLeftMargin{98. }%
\CSLRightInline{\textbf{Identification of glioblastoma~gene prognosis modules based on weighted gene co-expression network analysis}
\CSLBlock{Pengfei Xu, Jian Yang, Junhui Liu, Xue Yang, Jianming Liao, Fanen Yuan, Yang Xu, Baohui Liu, Qianxue Chen} \emph{BMC Medical Genomics} (2018-11-01) \url{https://doi.org/gfzt4j}
\CSLBlock{DOI: \href{https://doi.org/10.1186/s12920-018-0407-1}{10.1186/s12920-018-0407-1} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/30382873}{30382873} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6211550}{PMC6211550}}}

\leavevmode\vadjust pre{\hypertarget{ref-KNAOTAKP}{}}%
\CSLLeftMargin{99. }%
\CSLRightInline{\textbf{Boosting GWAS using biological networks: A study on susceptibility to familial breast cancer}
\CSLBlock{Héctor Climente-González, Christine Lonjou, Fabienne Lesueur, Dominique Stoppa-Lyonnet, Nadine Andrieu, Chloé-Agathe Azencott} \emph{PLOS Computational Biology} (2021-03-18) \url{https://doi.org/g7zdns}
\CSLBlock{DOI: \href{https://doi.org/10.1371/journal.pcbi.1008819}{10.1371/journal.pcbi.1008819} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/33735170}{33735170} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8009366}{PMC8009366}}}

\leavevmode\vadjust pre{\hypertarget{ref-i8Qjy4pw}{}}%
\CSLLeftMargin{100. }%
\CSLRightInline{\textbf{Adjusting for genetic confounders in transcriptome-wide association studies improves discovery of risk genes of complex traits}
\CSLBlock{Siming Zhao, Wesley Crouse, Sheng Qian, Kaixuan Luo, Matthew Stephens, Xin He} \emph{Nature Genetics} (2024-01-26) \url{https://doi.org/gtf7kb}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41588-023-01648-9}{10.1038/s41588-023-01648-9} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/38279041}{38279041} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC10864181}{PMC10864181}}}

\leavevmode\vadjust pre{\hypertarget{ref-1EIbaYHGq}{}}%
\CSLLeftMargin{101. }%
\CSLRightInline{\textbf{Graphics processing units in bioinformatics, computational biology and systems biology}
\CSLBlock{Marco S Nobile, Paolo Cazzaniga, Andrea Tangherloni, Daniela Besozzi} \emph{Briefings in Bioinformatics} (2016-07-08) \url{https://doi.org/g7zg6w}
\CSLBlock{DOI: \href{https://doi.org/10.1093/bib/bbw058}{10.1093/bib/bbw058} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/27402792}{27402792} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5862309}{PMC5862309}}}

\leavevmode\vadjust pre{\hypertarget{ref-4iZVPVHz}{}}%
\CSLLeftMargin{102. }%
\CSLRightInline{\textbf{CuPy: A NumPy-Compatible Library for NVIDIA GPU Calculations}
\CSLBlock{Ryosuke Okuta, Yuya Unno, Daisuke Nishino, Shohei Hido, Crissman Loomis} \emph{Proceedings of Workshop on Machine Learning Systems (LearningSys) in The Thirty-first Annual Conference on Neural Information Processing Systems (NIPS)} (2017)}

\leavevmode\vadjust pre{\hypertarget{ref-wPsNXL1Q}{}}%
\CSLLeftMargin{103. }%
\CSLRightInline{\textbf{RAPIDS \textbar{} GPU Accelerated Data Science}
\CSLBlock{RAPIDS \textbar{} GPU Accelerated Data Science} \url{https://rapids.ai/}}

\leavevmode\vadjust pre{\hypertarget{ref-1LexfAxj}{}}%
\CSLLeftMargin{104. }%
\CSLRightInline{\textbf{Array programming with NumPy}
\CSLBlock{Charles R Harris, KJarrod Millman, Stéfan J van der Walt, Ralf Gommers, Pauli Virtanen, David Cournapeau, Eric Wieser, Julian Taylor, Sebastian Berg, Nathaniel J Smith, \ldots{} Travis E Oliphant} \emph{Nature} (2020-09-16) \url{https://doi.org/ghbzf2}
\CSLBlock{DOI: \href{https://doi.org/10.1038/s41586-020-2649-2}{10.1038/s41586-020-2649-2} · PMID: \href{https://www.ncbi.nlm.nih.gov/pubmed/32939066}{32939066} · PMCID: \href{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7759461}{PMC7759461}}}

\leavevmode\vadjust pre{\hypertarget{ref-AujvwLp6}{}}%
\CSLLeftMargin{105. }%
\CSLRightInline{\textbf{Scikit-learn: Machine Learning in Python}
\CSLBlock{Fabian Pedregosa, Gaël Varoquaux, Alexandre Gramfort, Vincent Michel, Bertrand Thirion, Olivier Grisel, Mathieu Blondel, Andreas Müller, Joel Nothman, Gilles Louppe, \ldots{} Édouard Duchesnay} \emph{arXiv} (2012) \url{https://doi.org/gpptwn}
\CSLBlock{DOI: \href{https://doi.org/10.48550/arxiv.1201.0490}{10.48550/arxiv.1201.0490}}}

\leavevmode\vadjust pre{\hypertarget{ref-b5xmRoLZ}{}}%
\CSLLeftMargin{106. }%
\CSLRightInline{\textbf{PyCUDA and PyOpenCL: A scripting-based approach to GPU run-time code generation}
\CSLBlock{Andreas Klöckner, Nicolas Pinto, Yunsup Lee, Bryan Catanzaro, Paul Ivanov, Ahmed Fasih} \emph{Parallel Computing} (2012-03) \url{https://doi.org/c8hxzm}
\CSLBlock{DOI: \href{https://doi.org/10.1016/j.parco.2011.09.001}{10.1016/j.parco.2011.09.001}}}
